Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-05 paper.tex
conclusion added

%\def\del#1{{}}
\def\del#1{{\bf (DELETED TEXT)}}
%\def\C#1{#1}
\def\C#1{{\bf #1}}

@@

This yields Fornax ($\mvir=10^{14}\,\msun$) and Virgo
($\mvir=2.1\times10^{14}\,\msun$) \cite{} as the prime targets for DM
observations, Perseus ($\mvir=7.7\times10^{14}\,\msun$) for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma ($\mvir=1.4\times10^{15}\,\msun$) for comparison.

@@

@@

Some of the results (should discuss what we want to emphasize):
--) the flux from annihilating DM is boosted by about a factor 1000 in
 clusters
--) flat surface brightness profiles with substructures
--) Fornax is a great target for indirect DM studies because the relative
 low flux from CRps and high flux from DM
--) the IC upscattered photons from stars and dust are dominated in the
 relevant energy regimes by the continuum emission for our DM
 benchmark models
--) Fermi 3 year data will constrain the smallest mass of substructures
 using the benchmark models
--) difficult to detect DM without a Sommerfeld boost with Cherenkov
 telescopes since the boost from substructures is extended while the
 sensitivity of Cherenkov telescopes drops linearly with radius
 outside the point spread function (CHECK)
--) we rule out the leptophilic models in present form in 14 clusters,
 and limit the boost from Sommerfeld enhancement in the MW to less
 than 10
--) the tightest limits on DM and CRps from Fermi are in the 1-10 GeV
 regime, where Fornax, M49, and NGC4636 are the most constraining
 "clusters" for the DM and Virgo, Coma and Norma are very close to the
 predictions for the CRps
--) we provide a rather simple model to calculate the spatial and spectral
 distribution of dust and stars in cluster (might be used for
 galaxies? with different spatial profile)

>>>>

\def\del#1{{}}
%\def\del#1{{\bf (DELETED TEXT)}}
\def\C#1{#1}
%\def\C#1{{\bf #1}}

@@

This yields Fornax ($\mvir=10^{14}\,\msun$) and Virgo
($\mvir=2.1\times10^{14}\,\msun$) \del{\cite{}} as the prime targets for DM
observations, Perseus ($\mvir=7.7\times10^{14}\,\msun$) for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma ($\mvir=1.4\times10^{15}\,\msun$) for comparison.

@@

\clearpage

@@

In this paper, we study the possibility for detecting $\gamma$-ray emission in
galaxy clusters. We consider benchmark and leptophilic models of supersymmetric
dark matter as well as cosmic ray induced pion decay which is thought to
dominate the astrophysical signal from clusters. Once supersymmetric cold dark
matter decouples from the expanding universe it streams freely and smooths over
any potentially present smaller scales. This imprints as a characteristic
minimum halo mass $M_\mathrm{min}$ into the hierarchy of structure formation and
should still survive until the present time as substructure within larger
halos. It turns out that the boost due to substructures dominates over the
smooth component for halo masses $M>10^3 \msun$ and contributes a constant
luminosity for each decade in halo mass. The ratio of halo-to-minimum subhalo
mass $M_{200}/M_\mathrm{min}$ is maximized for the largest systems that had time
to collapse and virialize since the Big Bang -- galaxy clusters. The
corresponding substructure boost can reach values exceeding $10^3$ which should
make nearby clusters the brightest DM annihilation sources after the Milky May
center and hypothetical very close-by DM substructures. Since the substructures
are distributed according to an Einasto profile, the contribution to the
annihilation emission from substructures peaks around the virial radius. This
implies an almost flat surface brightness profile of the annihilation emission
which makes it necessary to have the entire cluster in the field-of-view to take
advantage of the entire substructure boost.  This property makes it difficult to
detect DM annihilation emission without an enhancement of the particle physics
cross section over its standard value of $\sigma v\sim 3\times 10^{-26}
~\mathrm{cm}^3~\mathrm{s}^{-1}$ with imaging air Cherenkov telescopes due to the
spatially extended boost from substructures of angular extend $\sim
0.5-1\deg$. In contrast, the sensitivity of Cherenkov telescopes drops linearly
with radius outside the point spread function $\sim0.1\deg$.

We identify Fornax to be the best cluster/group candidate for indirect DM
studies as it is expected to emit the brightest annihilation flux and more
importantly, has a comparably low CR induced $\gamma$-ray flux.  This in
principle should allow to detect or constrain some benchmark models that have
been chosen for the large $\gamma$-ray flux yield with three-year Fermi data
(or constrain the substructure boost factor). Other nearby bright sources of the
extended HIFLUCGS sample of nearby bright X-ray cluster/groups are M49,
Centaurus, and NGC~4636.  The non-detection of $\gamma$-ray emission by Fermi in
a total of 14 clusters/groups (the mentioned clusters as well as Virgo, Coma,
Hydra, and Norma among others) considerably constrains leptophilic models for DM
annihilation which were introduced to explain the increasing positron-to-lepton
ratio beyond 10 GeV. Assuming that the minimum DM substructure mass extends down
to an Earth mass, we can rule out these leptophilic models based on the
non-observation of $\gamma$-rays in any of the fore-mentioned
clusters/groups. Alternatively, assuming the leptophilic models to be correct,
we limit the minimum substructure mass to $4~\msun$ in Fornax. Assuming
universality of the DM model, this corresponds to a minimum substructure mass of
$10~\msun$ in the Milky Way due to the larger Sommerfeld boost factor in smaller
mass halos which have a lower circular velocity which enhances the particle
cross section.

In this work, we thoroughly study the spectral emission characteristics of the
DM emission. In general there are different radiation mechanism that emit in the
$\gamma$-ray regime, namely (i) continuum emission following the hadronization
of the annihilating neutralinos in benchmark models which leads to the
production of charged and neutral mesons which decay finally into electrons,
neutrinos, and gammas, and (ii) inverse Compton (IC) emission by the leptons in
the final state which up-scatter radiation fields from the cosmic microwave
background, dust, and starlight, and (iii) final state radiation or internal
bremsstrahlung which dominates the highest energy emission close to the rest
mass of the self-annihilating neutralinos (if present in the model under
consideration). We construct a rather simple semi-analytic model for the spatial
and spectral distribution of dust and starlight in clusters which can be used
for the annihilation emission in galaxies with a different spatial profile.  We
find that in benchmark models, the continuum emission dominates over the IC
emission of any available radiation field. In the leptophilic models, where the
continuum emission is absent, the IC up-scattering of the CMB dominate below 30
GeV. For energies $30~\mathrm{GeV} \lesssim E_\gamma \lesssim 300~\mathrm{GeV}$,
the IC up-scattering of the dust emission dominates and is followed by the final
state radiation for energies up to $m_\chi c^2 \sim 1$~TeV. The IC emission of
starlight remains always subdominant. We stress, that the inclusion of the dust
and starlight radiation fields are important in the cooling function for the
leptons and a failure to do so biases the inferred CMB IC flux high by a factor
of two.

We substantially improve the modelling of the expected $\gamma$-ray signal from
pion-decay resulting from hadronic CR interactions with gas protons. We employ
the semi-analytic model of the CR spectrum and spatial distribution, recently
developed by \citet{2010MNRAS.409..449P} that is based on cosmological
hydrodynamical simulations that self-consistently follow the evolution of CRs in
cosmic structure formation. Adopting the model for all clusters in the HIFLUGCS
sample of flux-limited X-ray clusters, we compare the expected $\gamma$-ray
emission to that inferred from using $L_\gamma$-cluster mass scaling
relation. We note that the predictions of our ``optimistic'' model are perfectly
in agreement with Fermi upper limits for all clusters.  The hint of a tension
that the scaling relation modeling showed for Norma \citep{2010ApJ...717L..71A}
could not be confirmed due to the apparently lower gas density of this cluster
relative to the average.  As expected, the semi-analytic model adds ``scatter''
to the scaling relation and biases the $\gamma$-ray flux high for prominent cool
core clusters by up to a factor of 3.5 relative to the expectation from the
scaling relation. This is due to the high target gas densities and associated CR
densities in those cool cores that shape the spatial emission characteristic to
be very similar to that observed in thermal X-rays. We caution that these
predictions only apply for the case of negligible CR transport such as streaming
and diffusion relative to the gas and might be considerably modified in those
cluster that do not show an extended diffuse radio(-mini) halo that might point
to a centrally concentrated CR distribution \citep{2011A&A...527A..99E}. The CR
spectrum $E^2 dN/dE$ shows the characteristic pion bump at energies around
1~GeV. It is expected to dominate the DM annihilation signal in the benchmark
models for all clusters/groups but Fornax. We conclude that it will be possible
to rule out the leptophilic model with eight-years of Fermi data and start to
constrain some benchmark DM models. Hence this should nicely complement direct
and accelerator searches for the nature of dark matter.

\del{
Some of the results (should discuss what we want to emphasize):
O) the flux from annihilating DM is boosted by about a factor 1000 in
 clusters
O) flat surface brightness profiles with substructures
O) Fornax is a great target for indirect DM studies because the relative
 low flux from CRps and high flux from DM
O) the IC upscattered photons from stars and dust are dominated in the
 relevant energy regimes by the continuum emission for our DM
 benchmark models
O) Fermi 3 year data will constrain the smallest mass of substructures
 using the benchmark models
O) difficult to detect DM without a Sommerfeld boost with Cherenkov
 telescopes since the boost from substructures is extended while the
 sensitivity of Cherenkov telescopes drops linearly with radius
 outside the point spread function (CHECK)
O) we rule out the leptophilic models in present form in 14 clusters,
 and limit the boost from Sommerfeld enhancement in the MW to less
 than 10
--) the tightest limits on DM and CRps from Fermi are in the 1-10 GeV
 regime, where Fornax, M49, and NGC4636 are the most constraining
 "clusters" for the DM and Virgo, Coma and Norma are very close to the
 predictions for the CRps
O) we provide a rather simple model to calculate the spatial and spectral
 distribution of dust and stars in cluster (might be used for
 galaxies? with different spatial profile)
}
